Layered Types -- A structured approach to arbitrarily combine Liquid Types

### 2023-01-25 paper/paper.tex
* Finished paper chapter on file handling use case

Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types. This forces us to combine all properties we want to statically check and verify into one big predicate. Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs. Additionally previously declared liquid types may not be reused when wanting to refine them further.

We therefore propose an implementation of \textit{Layered Liquid\todo{Just Layered or Layered Liquid?} Types}, that allows to define types as a conjunction of previously defined Liquid Types. These types are then checked iteratively to ensure compability between functions and arguments.

Assume we have a function \texttt{f} that takes as input a single liquid type $T^{in}$ refined by the predicates $p^{in}_1,\cdots,p^{in}_n$ and result type \texttt{out} and a variable \texttt{arg} of liquid type $T^{arg}$ refined by the predicates $p^{arg}_1,\cdots,p^{arg}_m$:

\begin{lstlisting}[mathescape=true]
	-- Liquid Type Definitions
	{-@ type $T^{in}$ t = { t | $p^{in}_1 \wedge ... \wedge p^{in}_n$  } @-}
	{-@ type $T^{arg}$ t = { t | $p^{arg}_1 \wedge ... \wedge p^{arg}_m$ } @-}

	f : $T^{in}$ -> out 
	arg : $T^{arg}$
\end{lstlisting}

When now calling \texttt{f} with \texttt{arg} as an argument, current implementations of liquid types will check for compability of the arguments in a manner similar to:

$$ \left( \bigwedge^{m}_{i=1} p^{arg}_i \right) \implies \left( \bigwedge^{n}_{j=1} p^{in}_j \right) $$

In our layered approach we rather propose that we can define separate layered types for each predicate and easily use their conjunction as the specific refined type for arguments and values:

\begin{lstlisting}[mathescape=true]
	-- Liquid Type Definitions for inputs to f
	{-@ type $T^{in}_1$ t = { t | $p^{in}_1$ } @-}
	{-@ type ... @-}
	{-@ type $T^{in}_n$ t = { t | $p^{in}_n$ } @-}

	-- Liquid Type Definitions for arg
	{-@ type $T^{arg}_1$ t = { t | $p^{arg}_1$ } @-}
	{-@ type ...  @-}
	{-@ type $T^{arg}_m$ t = { t | $p^{arg}_m$ } @-}

	f : {$T^{in}_1 \wedge ... \wedge T^{in}_n$} -> out 
	arg : {$T^{arg}_1 \wedge ... \wedge T^{arg}_m$}
\end{lstlisting}

In this layered approach predicates are not checked simultaneously but rather in an iterative way, which could intuitively rather be expressed as the (which is semantically equivalent to the version above):

$$ \bigwedge^{n}_{j=1} \left( \bigwedge^{m}_{i=1} p^{arg}_i \implies  p^{in}_j \right) $$

While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques. While the latter deal with the gradient between incomplete and complete information (in that case on types), our approach is rather concerned not with what information is available on the typing but rather how it is processed. While gradual approaches might still type-check with incomplete typing information our approach would fail to typecheck in such a scenario as it still requires complete information.

@@

\begin{figure}[ht!]
\begin{align*}
	id \in \texttt{Ident}\ 	&::=\ [a...zA...Z]+ \\
	v \in \texttt{Var}\ 	&::=\ id \\
	f \in \texttt{Fun}\ 	&::=\ id_1\ id_2\ \{\ s\ \}\\
	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e) \\
	\texttt{CustomExpr}\ 	&::=\ \texttt{String}\\
	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
	\texttt{Assign}\ 	&::=\ v\ :=\ e\ |\ v\ :=\ \texttt{CustomExpr}\ \\
	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ s\ \texttt{else}\ s\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ s\ |\ \ell\ |\ s_1\ s_2\ \\
	\ell \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
	\circleddot\ 		&::=\ \texttt{!=}\ |\ \texttt{==}\ |\ \texttt{<} \ |\ \texttt{>} \ |\ \texttt{<=} \ |\ \texttt{>=} \\
\end{align*}
\caption{Syntax of the simple functional language}
\label{fig:lang}
\end{figure}

@@

These layers can then be used to refine an identifier. In a very basic and minimal approach these layers can be used to annotate types to variables and functions. Additionally, one might add additional layers to a variable to ensure more properties. Each layer will be checked individually and independently. A more detailled approach of the architecture behind our layered approach will be described in section \ref{ssec:architecture}.

An example of a program written in our language can be seen in \ref{lst:fac_example}. The program defines a factorial function that is refined in two layers. The first layer \textit{base} is a simple type layer that defines that \texttt{fac} takes an \texttt{Int} and returns an \texttt{Int}. The second layer \textit{liquid} is a refinement layer that refines that the input should be greater or equal to zero and that the output will be greater than zero.

@@

The implementation of our language will be done in the \texttt{Python} programming language. Roughly speaking, the implementation will consist of three parts, which we will describe throughout this section:

@@

After initially parsing the AST, it will contain explicit nodes for each of the defined layers. As a first processing step we use a Transformer to collect these layers and remove the corresponding nodes from the AST. The implementation of each node has to be provided by the user as a separate Python module. Our language Interpreter will then dynamically load the corresponding module and use it to check the layer. To interact with our layered type checker correctly the module has to implement the following functions:

\begin{itemize}
	\item \texttt{depends\_on} -- This function returns a list of all layers that this layer depends on. This is used to ensure that the layers are checked in the correct order.
	\item \texttt{parse\_type} -- This function may be used to define and parse custom data types that are defined usinf the $CustomExpr$ rule in the grammar. \todo{Implementation details -- When and how is this function called?}
	\item \texttt{check\_layer} -- This function is used to check the layer. It is passed the AST of the program with all the annotated information from the previous layers. It should return a boolean value that indicates whether the layer is valid or not.
\end{itemize}

Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers. It will then topologically sort the layers and check them in the correct order. Once all layers have been checked successfully, the interpreter will execute the program.

Using this architecture, we provide a flexible framework that allows to easily extend our toy language with various type checking layers. Similar to passes in a compiler, the layers can be added and removed easily. Additionally, the layers can be easily combined to create more complex type checking layers.

\subsection{Examples}
\label{ssec:examples}

>>>>

% Current Liquid Type implementations limit us to only defining one liquid type for variables, arguments or return types. This forces us to combine all properties we want to statically check and verify into one big predicate. Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs. Additionally previously declared liquid types may not be reused when wanting to refine them further.
Current verification approaches usually focus on verifying a single property of a program at a time. However, in practice, one might needs to verify different properties of the program for different purposes. In current implementations, it is not trivially possible to combine these different verification passes. One specific examplefor this can be found when looking at current implemetnations of Liquid Types. In these, one can only define one liquid type for variables, arguments or return types. This forces us to combine all properties we want to statically check and verify into one joint predicate. Using such an approach has the drawback that all properties are checked simultaneously which might make it difficult for a developer to identify due to which property a type error occurs. Additionally previously declared liquid types may not be reused when wanting to refine them further.

With this paper, we therefore propose a new incremental verification approach that we will refer to as \textit{Layered Types}. The goal of this approach is it for a program to be possible to define individual layers for the verification process that are verified individually. By defining dependencies between those layers, it is possible to define a verification order that ensures that all layers are verified in the correct order. This allows us to define different properties of a program that can be verified independently from each other.

While in this paper, we will focus on specific Use-Cases for Liquid Types, we believe that the proposed approach can be applied to other verification techniques as well. For example, we believe that the proposed approach can be extended to verify all sorts of different properties and typesystems such as e.g. \textit{Linear Types} \cite{linear_types} or \textit{Dependent Types} \cite{dependent_types}.

While on first glance this approach could be described as an approach to \textit{Gradual Verification}, we will explicitly refer to our approach as \textit{Layered} to distinguish it from \textit{Gradual Verification} techniques. While the latter deal with the gradient between incomplete and complete information (in that case on types), our approach is rather concerned not with what information is available on the typing but rather how it is processed. While gradual approaches might still type-check with incomplete typing information our approach would fail to typecheck in such a scenario as it still requires complete information. However, we believe it would be possible to embed certain gradual verification techniques as independent layers in our approach.

@@

\begin{figure}[ht!]
\begin{align*}
	id \in \texttt{Ident}\ 	&::=\ [a...zA...Z]+ \\
	v \in \texttt{Var}\ 	&::=\ id \\
	f \in \texttt{Fun}\ 	&::=\ id_1\ ...\ id_n\ \{\ s\ \}\\
	e \in \texttt{Expr}\ 	&::=\ v\ |\ c\ |\ (\ e\ \circledplus\ e\ )\ |\ (e\ \circleddot\ e)\ |\ f(e)\ |\ f(e_1,\ ...,\ e_n)  \\
	\texttt{CustomExpr}\ 	&::=\ \texttt{String}\\
	c \in \texttt{Const}\ 	&::=\ \texttt{True}\ |\ \texttt{False}\ |\ \mathbb{Z} \\
	\texttt{Assign}\ 	&::=\ v\ :=\ e\ |\ v\ :=\ \texttt{CustomExpr}\ \\
	s \in \texttt{Stmt}\ 	&::=\ \texttt{Assign}\ |\ \texttt{if}\ v\ \texttt{then}\ \lbrace\ s_1\ \rbrace\ \texttt{else}\ \lbrace\ s_2\ \rbrace\ |\ \texttt{let}\ v\ :=\ e\ \texttt{in}\ \lbrace\ s\ \rbrace\ |\ \ell\ |\ s_1\ s_2\ \\
	\ell \in \texttt{Layer}\ 	&::=\ id_1::id_2::\texttt{String} \\
	\circledplus\ 		&::=\ \texttt{+}\ |\ \texttt{-} \\
	\circleddot\ 		&::=\ \texttt{!=}\ |\ \texttt{==}\ |\ \texttt{<} \ |\ \texttt{>} \ |\ \texttt{<=} \ |\ \texttt{>=} \\
\end{align*}
\caption{Syntax of the simple functional language}
\label{fig:lang}
\end{figure}

@@

These layers can then be used to define properties. Depending on the Layers' implementation, these properties can apply to identifiers or can be used to define properties of the type system the layer defines. In a very basic and minimal approach these layers can be used to annotate types to variables and functions. Additionally, one might add additional layers to a variable to ensure more properties. Each layer will be checked individually and independently. A more detailled approach of the architecture behind our layered approach will be described in section \ref{ssec:architecture}.

An example of a program written in our language can be seen in \ref{lst:fac_example}. The program defines a factorial function that is refined in two layers. The first layer \textit{base} is a simple type layer that defines that \texttt{fac} takes an \texttt{Int} and returns an \texttt{Int}. The second layer \textit{liquid} is a refinement layer that refines that the input should be greater or equal to zero and that the output will be greater than zero. The implementations of the \textit{base} and \textit{liquid} layer are not part of our language and have to be implemented independently.

@@

The implementation of our language will be done in the \texttt{Python} programming language. Generally the implementation can be divided into three individual parts, which we will describe throughout this section:

@@

After parsing the AST there are a few simple checks that are run by an initial interpreter pass. Namely, our framework will check that variables are assigned to at least once before they are used in another expression, it will make sure that \texttt{let} does not redeclare an already existing variable. Additionally, function names cannot be defined twice and the framework will also check that functons that are called have either been defined previously in the source file or alternatively are defined externally in Python.

After this initial check has passed, the main part of our work, the layered type checker, will start processing the layers. A "Layer" here refers to a custom implementation that has to be defined in a separate Python file by the user. Based on the defined layer name, our framework will dynamically load the python file as an independent module and will use it to run the type-checker for this layer. Our framework supports the definition of the following functions for each layer:

\begin{itemize}
	\item \texttt{depends\_on} -- This function returns a list of strings. Each string represents a name of a layer that this layer depends on and thus has to be checked before this layer can be run. This is used to ensure that the layers are checked in the correct order.
	\item \texttt{run\_before} -- This function is similar to \texttt{depends\_on} but returns a list of layers that must not run before this layer has been checked corretly. Together with the \texttt{depends\_on} function this allows to flexibly insert layers into any step of the type-checking process.
	\item \texttt{parse\_type} -- This function may be used to define and parse custom data types that are defined usinf the $CustomExpr$ rule in the grammar. \todo[inline]{This is a functionality we discussed in the beginning, that would be nice. However for our current use cases we do not really need it. Should we keep it in the framework or just omit it?}
	\item \texttt{typecheck} -- This function is used to check the layer. It is passed the AST of the program with all the annotated information from the previous layers. It should return the annotated tree with the new layer information added.
\end{itemize}

Using the \texttt{depends\_on} function, the layered type checker will first build a dependency graph of all layers. By using the dependency graph we can calculate the topological order of layers that can be checked in the correct order. Additionally this also allows to check independent layers in parallel. Only if all layers have been checked successfully, the program will be executed.

In order to communicate annotated information between the different layers, we replace the \texttt{Tree} \Lark{} class with our own \texttt{AnnotatedTree} class. This class allows us to store a dictionary of annotations for each node in the tree. The annotations are stored as a dictionary of dictionaries. The first key of the dictionary is the name of the layer, the second key is the name of the annotation. This allows us to store multiple annotations for each layer. The \texttt{AnnotatedTree} class also provides a simple interface to access and add annotations which can be used by developers implementing new layers:

\begin{itemize}
	\item \texttt{get\_layer\_annotation(layer\_id, identifier, key)} -- This function returns the annotation with the given key for the given identifier in the given layer. If no annotation is found, it returns \texttt{None}.
	\item \texttt{get\_all\_layer\_annotations(layer\_id, identifier)} -- This function returns a dictionary of all annotations for the given identifier in the given layer. If no annotations are found, it returns an empty dictionary.
	\item \texttt{add\_layer\_annotation(layer\_id, identifier, key, value)} -- This function adds a annotation with the given key for the given identifier in the given layer to the given value.
	\item \texttt{update\_layer\_annotations(layer\_id, identifier, values)} -- This updates the layer annotations for the given identifier in the given layer with the given dictionary of values.
\end{itemize}

Using this architecture, we provide a flexible framework that allows to easily extend our toy language with different layers that can be used to implement different type systems. Similar to passes in a compiler, the layers can be added and removed easily. Additionally, the layers can be easily combined to create more complex type systems.

\section{Example Use-Cases}
\label{ssec:use_cases}
For the scope of this paper, we have developed two exemplary use-cases for our framework that demonstrates our layered verification approach. In the first use-case, we defined a simple program that uses multiple different, partly independent type systems. For our second example, we use our framework to present a more flexible approach to using Liquid Types.

\subsection{Use Case 1 -- File Processing}
\label{sssec:use_case_1}

In our first use case we implemented a small program that processes the contents of a text file. Througout this section we will describe the program and the implemented layers required to realize this program. Lastly we will discuss the findings and benefits we gained from using our layered verification approach.

\subsubsection{Description}
In our first use-case we design a small program that reads contents from a file and performs certain operations based on that. The base code for our example can be found in Listing \ref{lst:uc1_code}. In our example, the file to be read will contain one number per each line. Depending on whether the number is positive or negative, we add the number as is or the square of the number to the resulting value.

\begin{lstlisting}
processLine lines idx length {
    end := idx == length

    if end then {
        0
    } else {

        curLine := getLine(lines, idx)

        number := strToInt(curLine)

        greaterThanZero := number > 0

        if greaterThanZero then {
            number + processLine(lines, idx + 1, length)
        } else {
            (number * number) + processLine(lines, idx + 1, length)
        }

    }
}

processLines lines {
    length := len(lines)

    idx := 0
    processLine(lines, idx, length)
}

fileName := "numbers.data"
fileHandle := openFile(fileName)
lines := readLines(fileHandle)
closeFile(fileHandle)
total := processLines(lines)

print(total)
total

\end{lstlisting}

All functions that have not been defined directly \ref{lst:uc1_code} are provided in a separate, external python file. As our language on itself does not include a type system and there are no layer annotations in the code above, no type checking will be performed on the code as it is presented.

However even this simple example could already introduce several errors that could be prevented by introducing some static analysis. First of all, it does not provide any information about the variable and function type and thus, we cannot ensure that functions are called with arguments of the correct types which might lead to runtime errors.

Additionally we our code interacts with the file it operates on through a file descriptor. TThis file descriptor can be viewed as a resource that might have an implicit or explicit internal state. In a very simple approach this file descriptor could be in the states "Open" or "Closed". In order to read from it, it should be in an "Open" state and additionally we should ensure that by the end of our program the resource has been properly closed again.

In order to improve our example code and allow static analysis, we therefore will introduce three layers:
\begin{enumerate}
	\item A \textit{Type} Layer -- This layer will be used to simply define variable and function types
	\item The \textit{Typechecking} Layer -- This layer will allow to flexibly define a type system and then use the information provided by the \textit{Type} layer to check if the program is well-typed. It therefore depends on the \textit{Type} layer to provide the type information.
	\item A \textit{State} Layer -- This layer will be used to define states on variables. Additionally it can model state transitions. We will use it to track the state of the file descriptor. This layer operates completely independent from the \textit{Type} and \textit{Typechecking} layer.
\end{enumerate}

\subsubsection{Implemented Layers}
In the following sections we will explain in more detail how the layers are implemented and what Syntax they use.

\paragraph{Types Layer}
The \textit{Types} layer is the first layer that is checked. It is used to define the types of variables and functions. To define the type of a variable we use the \texttt{Layer} syntax of our language. A layer definition for the \textit{Types} layer looks as follows:

\begin{lstlisting}
identifier :: types :: Type
\end{lstlisting}

The \texttt{identifier} is the name of the variable or function for which we want to define the type. The \texttt{types} keyword is used to associate this information with the correct layer. The \texttt{Type} is the desired type information to be annotated. There are no restrictions on how a \texttt{Type} should look, as the types layer simply annotates the given string into the current node and does not further process it. For example to define the type of a variable \texttt{foo} as \texttt{int} we would write:

\begin{lstlisting}
foo :: types :: int
\end{lstlisting}

In order to define the type of a function, there is a slight adaption to this syntax. While the first two parts are identical to the variable definition, we now need to provide type information for each argument of the function. This is done by using the \texttt{Type} syntax for each argument and then separating them with an arrow symbol \texttt{->}. The return type of the function is then defined by the last \texttt{Type}. If a function does not take an argument the defintion simply starts with the arrow symbol. The following code piece contains an example for a function \texttt{f} that takes two arguments of type \texttt{int} and returns a value of type \texttt{bool}, and a function \texttt{g} that does not take any arguments and returns a value of type \texttt{string}:

\begin{lstlisting}
f :: types :: int -> int -> bool
g :: types :: -> string
\end{lstlisting}

It is important to note that the \textit{Types} layer does not perform any type checking. It simply annotates the given type information into the AST. The \textit{Types} layer only checks for the following inconsistencies:

\begin{itemize}
	\item The type for a variable can only be defined once in the current scope.
	\item When encountering a function definition, the type for it needs to be defined before the function signature. Additionally the number of arguments in the function signature needs to match the number of arguments in the type definition.
\end{itemize}

Any further type checking is performed by the \textit{Typechecking} layer.

\paragraph{Typechecking Layer}

The \textit{Typechecking} layer is used to define a type system and then use the information provided by the \textit{Types} layer to check if the program is well-typed. It therefore depends on the \textit{Types} layer to provide the type information. In our implementation of the type checker, we do not impose any restrictions on the specific type system. We rather provide a generic type checking implementation that allows the user to flexibly define their own types and subtype relationships. However, since our language implicitly natively requires specific expressions to be of a number or boolean type, there are implicitly part of our type system.

To define the specific type system, we use the \texttt{Layer} syntax of our language. However, instead of defining layers for identifiers as in the \textit{Types} layer, we reserve a selected set of keywords to define the type system. Depending on the definition provided by the user, the type checker will internally build a type graph that is used to perform the type checking.

The following keywords are reserved and can be used to define parts of the type system:

\begin{itemize}
	\item \texttt{logicalTypes} -- This keyword is used to define a set of logical types. Multiple types can be defined by separating them with a space.
	\item \texttt{numTypes} -- This keyword is used to define a set of numeric types. Multiple types can be defined by separating them with a space.
	\item \texttt{subtype} -- This keyword is used to define a subtype relationship between two or more types. The syntax for a subtype relationship is \texttt{<type1> <: <type2> <: ... <: <typeN>}. The subtype relationship is then defined by the order of the types. The first type is the supertype and the following type is a subtype of the previous type.
\end{itemize}

The following code piece is a short demonstration of a type system containing different logical types, numeric types and subtype relationships:

\begin{lstlisting}
logicalTypes :: typecheck :: bool
numTypes :: typecheck :: int float
subtype :: typecheck :: float <: int
\end{lstlisting}

The type system defined by the user will be used to perform the typechecking. To perform this typecheck the type checker accesses the annotations provided by the \textit{Types} layer. For the case that a type is neither a numeric or logical type and it does not have any subtype relationships, it is not required to explicitly define it in the type system. The type checker will then simply assume the existance of such a type that is not a subtype of any other type nor does it have any subtypes.

Our type checker will perform the following checks:

\begin{itemize}
	\item The type of each variable needs to be defined.
	\item When assigning a value to a variable, the type of the value needs to be a subtype of the type of the variable.
	\item When performing a logical operations (e.g. \texttt{\&\&}, \texttt{||}) the type of the operands needs to be a logical type. The result of the operation will be a logical type.
	\item When performing a numeric operation (e.g. \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}) the type of the operands needs to be a numeric type. The result of the operation will be the most general numeric type of the operands.
	\item When performing a comparison operation (e.g. \texttt{==}, \texttt{!=}, \texttt{<}, \texttt{>}, \texttt{<=}, \texttt{>=}) the type of the operands needs to be compatible. This means that either the right hand side is a subtype of the left hand side or the left hand side is a subtype of the right hand side. The result of such an operation is always a boolean value.
	\item In an \texttt{if} statement the condition needs to be a logical type.
	\item When performing a function call, the number of arguments needs to match the number of arguments in the function definition. Additionally the type of each argument needs to be a subtype of the type of the corresponding argument in the function definition.
\end{itemize}

Since we require the type of each variable to be defined, this means we require full information about the types of all variables. We do not perform any type inference. This is a deliberate design decision, as we want to keep the language as simple as possible. However, by the design of our language, it would be possible to develop an additional layer that runs after the \textit{Types} layer but before the \textit{Typechecking} layer that performs a type inference based on the incomplete information provided.

\paragraph{State Layer}

The \textit{State} layer is used to define states on variables. Similarly, we can use it to check if a variable has a certain state and define state transitions. This can for example be used to track the state of resources that are used by a program. In our implementation, the state of a variable is completely independent of its type, therefore it does not require any dependencies on the \textit{Types} layer.

The state of a variable is defined by a set of strings. Generally there are two ways to check the state of a variable or transition it's state to a new one. Firstly, we can explicitly do this by using a layer statement of the language. Secondly, we can also impose state requirements on arguments of function calls. Then when passing a variable to a function, the layer will check if the variable has the required state and apply the state transition if there are any.

Similar to the \textit{Types} layer, the \textit{State} layer uses the \texttt{Layer} syntax of the language to define the state of a variable. The first part of the layer statement defines the variable or function name to define a state on, while the second part is simply the layer identifier \texttt{state}. The last part of the layer definition defines the state.

A state definition is enclosed by curly brackets and can contain the following parts:

\begin{itemize}
	\item A state \textit{requirement} -- Imposes a restrition on the current state the variable needs to have. This is simply a string that needs to be part of the current state of the variable.
	\item A state \textit{transition} -- Defines a transition to a new state. Generally a state transition has the form \texttt{<oldState> => <newState>}. However, either the old or new state may be omitted. When omitting the old state, this simply means that something is added to the current state of the variable unconditionally. When omitting the new state, this means that the corresponding state string will be removed from the set of states.
\end{itemize}

For our use-case, we want to track the state of the file descriptor. Specifically, we want to make sure, that the file handle is open before we can read from it. Additionally, we do not want to read from a file handle again once it has already been read. Therefore, we define the states \texttt{Open}, \texttt{Consumed} and \texttt{Closed} and define the following state transitions for the functions that interact with the file descriptor:

\begin{lstlisting}
createFD :: state :: {} -> { Closed }
openFile :: state :: {Closed => Open} -> {}
readLines :: state :: {Open => Consumed} -> {}
closeFile :: state :: {Consumed => Closed} -> {}
\end{lstlisting}

\subsection{Use Case 2 -- Liquid Types}
\label{sssec:use_case_2}
